\section{Numerical experiments details} \label{2506:app:numerical-details}
All the codes used to run the experiments are available at \href{https://github.com/IdePHICS/Sequence-Single-Index}{this GitHub repository}; where details for reproducing figures are not available in the paper, we provide the code to reproduce them in the repository.

The experiments run on a Mac Studio M2 Ultra, within at most few hours for the largest ones. The code is written in Python, using the libraries \texttt{numpy}, \texttt{scipy}, \texttt{torch} and \texttt{matplotlib}. \texttt{hydra} is used to manage the configuration files.


\subsection{Details on the integration method of squared loss}
All the plot of population loss we showed are a numerical integration of the loss function. As showed in Section~\ref{2506:sec:setting}, the population loss is given by a multivariate Gaussian integral of \(2L\) dimensions, where the mean and the covariance are determined by the sufficient statistics. The integral can't be computed analytically, so we use a custom numerical procedure based on the Gauss-Hermite quadrature.

Let \(f\colon \mathbb{R}^{2L} \to \mathbb{R}\) be a function of \(2L\) variables to be integrated, and let \(\mu\in\mathbb{R}^{2L}\) and \(\Sigma\in\mathbb{R}^{2L\times 2L}\) be the mean and covariance of the multivariate Gaussian distribution. The integral we want to compute is
\begin{equation}
    I=\int_{\mathbb{R}^{2L}} f(x) \frac{1}{(2\pi)^{L}\sqrt{\det \Sigma}} \exp\left(-\frac{1}{2} (x - \mu)^\top \Sigma^{-1} (x - \mu)\right) dx
\end{equation}
The numerical procedure is as follows:
\begin{enumerate}
    \item Compute the 1D Gauss-Hermite nodes and weights for \(N_\mathrm{int}\) points
          \[
            \{x_i\}_{i=1}^{N_\mathrm{int}} \text{ and } \{w_i\}_{i=1}^{N_\mathrm{int}} 
          \]
          where \(x_i\) are the nodes given by the roots of the Hermite polynomial \(H_{N_\mathrm{int}}(x)\) and \(w_i\) are the corresponding weights, computed as
            \[
                w_i = \frac{2^{N_\mathrm{int}+1} N_\mathrm{int}! \sqrt{\pi}}{H_{N_\mathrm{int}}'(x_i)^2}
            \]
    \item Compute the \(2L\) dimensional nodes and weights by taking the Kronecker product of the 1D nodes and weights
          \[
            \{X_i\}_{i=1}^{N_\mathrm{int}^{2L}} = \bigotimes_{l=1}^{2L} \{x_i\}_{i=1}^{N_\mathrm{int}} \text{ and } \{W_i\}_{i=1}^{N_\mathrm{int}^{2L}} = \bigotimes_{l=1}^{2L} \{w_i\}_{i=1}^{N_\mathrm{int}}
          \]
    \item Let \(T\) be the Cholesky decomposition of \(\Sigma\), i.e. \(\Sigma = T^\top T\). We can then change the variable to \(y = T^{-1}(x - \mu)\) and compute the integral as
          \[\begin{split}
            I=&\int_{\mathbb{R}^{2L}} f(x) \frac{1}{(2\pi)^{L}\sqrt{\det \Sigma}} \exp\left(-\frac{1}{2} (x - \mu)^\top \Sigma^{-1} (x - \mu)\right) dx \\=& \int_{\mathbb{R}^{2L}} f(Ty + \mu) \frac{1}{(2\pi)^{L}} \exp\left(-\frac{1}{2} y^\top y\right) dy
          \end{split}\]
    \item The integral can then be approximated, as
          \[
            I \approx \sum_{i=1}^{N_\mathrm{int}^{2L}} \left(\prod_{l=1}^{2L}W_{i,l}\right) f(TX_i + \mu)
          \]
\end{enumerate}
The precision of the integral is obviously regulated by the number of points \(N_\mathrm{int}\) we use. In our experiments, we used \(N_\mathrm{int} = 17\), while for the phase diagram in Figure~\ref{2506:fig:phase-diagram-positional-semantic} we used \(N_\mathrm{int} = 19\). Some effects of this integration error are visible in Figure~\ref{2506:fig:app:phase-diagram-yellow-region}.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
