\chapter[Introduction]{Appendices to the Introduction}
\label{chap:appendix-introduction}

\section{The single-index overlap dynamics on the sphere}
\label{app:intro:m-ode-sphere}

This appendix supplies the derivation deferred from Section~\ref{sec:intro:exponents}. Throughout we are in the single-index, single-neuron setting with the read-out fixed at \(a = 1\) and the weights confined to the sphere \(\norm{\vec{w}} = \sqrt{d}\), so the self-overlap is frozen at \(q = 1\) and the pre-activations are jointly Gaussian, \((\lambda, \lambda^\star) \sim \mathcal{N}(0, \Omega)\) with \(\Omega = \left(\begin{smallmatrix} 1 & m \\ m & 1\end{smallmatrix}\right)\). Our goal is the compact overlap equation
\begin{equation} \label{app:intro:eq:target}
    \dod{m}{t} = (1 - m^2)\,C'(m), \qquad C(m) = \mathbb{E}_{(\lambda, \lambda^\star)}\!\left[ \sigma(\lambda)\,h^\star(\lambda^\star) \right],
\end{equation}
which we obtain in two complementary ways: directly from the discrete projected updates, in their gradient-flow limit on the sphere, which makes the origin of the factor \((1 - m^2)\) transparent; and by reducing the spherically projected Saad \& Solla drift~\eqref{eq:intro:m-ode-sphere}, which establishes the equivalence claimed in the main text.

\subsection{Derivation as a spherical gradient flow}
\label{app:intro:gradient-flow}

We start from the projected update of the main text: at step \(\tau\), on a fresh sample \((\vec{x}, y)\),
\begin{equation} \label{app:intro:eq:proj-update}
    \vec{w}^{\tau+1} = \sqrt{d}\,\frac{\vec{w}^{\tau} - \gamma\,\nabla_{\vec{w}}\ell}{\norm{\vec{w}^{\tau} - \gamma\,\nabla_{\vec{w}}\ell}}, \qquad \nabla_{\vec{w}}\ell = \big(\sigma(\lambda) - y\big)\,\sigma'(\lambda)\,\frac{\vec{x}}{\sqrt{d}}.
\end{equation}
Renormalizing onto the sphere makes the increment tangential to leading order in the step size,
\begin{equation} \label{app:intro:eq:tangent-step}
    \vec{w}^{\tau+1} - \vec{w}^{\tau} = -\gamma\,P_{\vec{w}}^{\perp}\,\nabla_{\vec{w}}\ell + O(\gamma^2), \qquad P_{\vec{w}}^{\perp} = I - \frac{\vec{w}\vec{w}^\top}{d},
\end{equation}
the discarded \(O(\gamma^2)\) carrying the curvature of the retraction together with the gradient fluctuations---the \emph{variance term} \(\Psi\), one power of \(\gamma\) down and neglected throughout, as in the main text. Since the sample is fresh the gradient is unbiased for the population risk, \(\mathbb{E}[\nabla_{\vec{w}}\ell] = \nabla_{\vec{w}}\mathcal{R}\), and---the projector depending on \(\vec{w}\) alone---
\begin{equation}
    \mathbb{E}[\vec{w}^{\tau+1} - \vec{w}^{\tau}] = -\gamma\,P_{\vec{w}}^{\perp}\,\nabla_{\vec{w}}\mathcal{R} + O(\gamma^2).
\end{equation}
On the sphere \(q = 1\) is frozen, so \(\mathcal{R}\) depends on the weights only through \(m = \sfrac{\langle \vec{w}, \vec{w}^\star\rangle}{d}\), and \(\mathcal{R}(m) = \mathrm{const} - C(m)\) as in Equation~\eqref{eq:intro:correlation-potential}. The chain rule gives the radial gradient \(\nabla_{\vec{w}}\mathcal{R} = \sfrac{\mathcal{R}'(m)}{d}\,\vec{w}^\star\); projecting it leaves \(P_{\vec{w}}^{\perp}\vec{w}^\star = \vec{w}^\star - m\,\vec{w}\), and reading off the overlap increment \(\mathbb{E}[\Delta m] = \sfrac1d\,\langle \vec{w}^\star,\,\mathbb{E}[\Delta \vec{w}]\rangle\),
\begin{equation}
    \mathbb{E}[\Delta m] = -\frac{\gamma}{d^2}\,\mathcal{R}'(m)\,\big(\norm{\vec{w}^\star}^2 - m\,\langle \vec{w}^\star, \vec{w}\rangle\big) = -\frac{\gamma}{d}\,\mathcal{R}'(m)\,(1 - m^2),
\end{equation}
using \(\norm{\vec{w}^\star}^2 = d\) and \(\langle \vec{w}^\star, \vec{w}\rangle = d\,m\). The Saad \& Solla time scaling now makes the dimension factor meaningful: measuring time as \(t = \tau\,\sfrac{\gamma}{d}\),
\addnote{In single-index setting the time scaling is \(t = \tau\,\sfrac{\gamma}{d}\) because \(p=1\).}
so that one sample advances \(t\) by \(\delta t = \sfrac{\gamma}{d}\), the overlap concentrates in the high-dimensional limit on the deterministic trajectory \(\dod{m}{t} = \mathbb{E}[\Delta m]/\delta t\). With \(\mathcal{R}'(m) = -C'(m)\) this is
\begin{equation}
    \dod{m}{t} = -(1 - m^2)\,\mathcal{R}'(m) = (1 - m^2)\,C'(m),
\end{equation}
Equation~\eqref{app:intro:eq:target}. The factor \(\sfrac{\gamma}{d}\) is thus the time step itself---not a stray constant---and \((1 - m^2)\) is the spherical Jacobian \(\sin^2\theta\) for \(m = \cos\theta\).

\subsection{Equivalence with the Saad \& Solla equation}
\label{app:intro:ss-equivalence}

The spherically projected Saad \& Solla drift~\eqref{eq:intro:m-ode-sphere} reads
\begin{equation}
    \dod{m}{t} = \mathbb{E}\!\left[(h^\star(\lambda^\star) - \sigma(\lambda))\,\sigma'(\lambda)\,\lambda^\star\right] - m\,\mathbb{E}\!\left[(h^\star(\lambda^\star) - \sigma(\lambda))\,\sigma'(\lambda)\,\lambda\right].
\end{equation}
For the bivariate Gaussian above, Stein's identities read, for any regular \(G(\lambda, \lambda^\star)\),
\begin{equation} \label{app:intro:eq:stein}
    \mathbb{E}[\lambda\, G] = \mathbb{E}[\partial_\lambda G] + m\,\mathbb{E}[\partial_{\lambda^\star} G], \qquad
    \mathbb{E}[\lambda^\star G] = m\,\mathbb{E}[\partial_\lambda G] + \mathbb{E}[\partial_{\lambda^\star} G],
\end{equation}
since \(\operatorname{Var}\lambda = \operatorname{Var}\lambda^\star = 1\) and \(\operatorname{Cov}(\lambda, \lambda^\star) = m\). Applying them to \(G = (h^\star(\lambda^\star) - \sigma(\lambda))\,\sigma'(\lambda)\), whose partial derivatives are \(\partial_\lambda G = (h^\star - \sigma)\,\sigma'' - (\sigma')^2\) and \(\partial_{\lambda^\star} G = {h^\star}'\,\sigma'\) (arguments suppressed), the two terms of the drift become
\begin{align}
    \mathbb{E}[\lambda^\star G] &= m\,\mathbb{E}\!\left[(h^\star - \sigma)\sigma'' - (\sigma')^2\right] + \mathbb{E}\!\left[{h^\star}'\sigma'\right], \\
    m\,\mathbb{E}[\lambda\, G] &= m\,\mathbb{E}\!\left[(h^\star - \sigma)\sigma'' - (\sigma')^2\right] + m^2\,\mathbb{E}\!\left[{h^\star}'\sigma'\right].
\end{align}
Subtracting, the self-interaction terms cancel identically and only the cross term survives,
\begin{equation}
    \dod{m}{t} = \mathbb{E}[\lambda^\star G] - m\,\mathbb{E}[\lambda\, G] = (1 - m^2)\,\mathbb{E}\!\left[{h^\star}'(\lambda^\star)\,\sigma'(\lambda)\right].
\end{equation}
Price's theorem---which states that, for jointly Gaussian \((\lambda, \lambda^\star)\) of unit variance and correlation \(m\),
\begin{equation} \label{app:intro:eq:price}
    C'(m) = \dod{}{m}\,\mathbb{E}[\sigma(\lambda)\,h^\star(\lambda^\star)] = \mathbb{E}[\sigma'(\lambda)\,{h^\star}'(\lambda^\star)]
\end{equation}
---then identifies this expectation with \(C'(m)\) and recovers Equation~\eqref{app:intro:eq:target}, establishing the equivalence.

\section{Validity window of the deterministic description}
\label{app:intro:validity-window}

The order-parameter ODEs of Equation~\eqref{eq:intro:ss-odes} rest on a
non-asymptotic guarantee, Theorem~\ref{thm:intro:ss-convergence}: over any fixed
horizon \([0, T]\) the stochastic overlap process stays within
\(\sfrac{C(T)}{\sqrt{d}}\) of the deterministic trajectory. The horizon enters
only through the constant \(C(T)\), and it is worth making its growth explicit.
Collecting the order parameters into a single vector \(\Omega\), the sharper form
established in Chapter~\ref{chap:paper-2302} (Theorem~\ref{2302:th:conv_eps}), due
to \citet{veiga2022phase}, reads, with step size \(\delta t = \sfrac{\gamma}{pd}\)
and continuous time \(t = \nu\,\delta t\),
\begin{equation} \label{app:intro:eq:veiga-bound}
    \lVert \Omega^{\nu} - \Omega(t) \rVert_\infty
    \;\leq\; e^{C t}\,\sqrt{\frac{\gamma}{pd}}, \qquad C > 0 .
\end{equation}
Since \(\sfrac{\gamma}{p} = \mathcal{O}(1)\) we have \(\sqrt{\sfrac{\gamma}{pd}}
\asymp d^{-1/2}\), so \eqref{app:intro:eq:veiga-bound} is the bound
\(\sfrac{C(T)}{\sqrt{d}}\) with the horizon dependence made visible:
\(C(T) \asymp e^{CT}\), the guarantee degrading \emph{exponentially} in time. It is
this bound that we invoked when we called the deterministic description ``valid
beyond the high-dimensional limit'' in Section~\ref{sec:intro:beyond}: the
right-hand side vanishes whenever \(\sfrac{\gamma}{pd}\to 0^{+}\), however that
limit is reached. What that statement leaves implicit, and what we make precise
here, is that the exponential prefactor confines the guarantee to a finite time
horizon.

\paragraph{The horizon.}
Fix a tolerance \(\eta \in (0,1)\) below which we regard the deterministic
trajectory as tracking the true process. The right-hand side of
\eqref{app:intro:eq:veiga-bound} stays under \(\eta\) exactly while
\begin{equation} \label{app:intro:eq:window}
    e^{Ct}\sqrt{\frac{\gamma}{pd}} \leq \eta
    \quad\Longleftrightarrow\quad
    t \;\leq\; t_\star \coloneqq \frac{1}{2C}\log\frac{pd}{\gamma}
        - \frac{1}{C}\log\frac{1}{\eta}
    \;=\; \frac{1}{2C}\log d + \mathcal{O}(1),
\end{equation}
again using \(\sfrac{\gamma}{p} = \mathcal{O}(1)\). The deterministic ODE is
therefore controlled only up to \(t_\star = \mathcal{O}(\log d)\); past it, the
error bar \eqref{app:intro:eq:veiga-bound} exceeds any fixed threshold and the
description guarantees nothing.

\paragraph{Where this leaves the escape times.}
Comparing \(t_\star\) with the information-exponent escape times of
Equation~\eqref{eq:intro:info-scaling} shows precisely which part of the dynamics
the deterministic ODE captures:
\begin{itemize}
    \item \(\ell = 1\): \(t_\text{escape} \asymp 1 \ll t_\star\); the escape
    happens well inside the window and the deterministic description is exact
    throughout.
    \item \(\ell = 2\): \(t_\text{escape} \asymp \log d\), the \emph{same} order
    as \(t_\star \asymp \sfrac{1}{2C}\log d\). The escape sits at the very edge of
    the guarantee: the deterministic drift still fixes the correct scaling, but
    whether the trajectory is resolved all the way out depends on the constant
    \(C\).
    \item \(\ell \geq 3\): \(t_\text{escape} \asymp d^{\ell/2 - 1}\), polynomially
    larger than \(t_\star\). Here \(e^{C t_\text{escape}} = e^{C\,d^{\ell/2-1}}\)
    is super-polynomially large long before escape, and the bound
    \eqref{app:intro:eq:veiga-bound} is \emph{vacuous} over the entire search
    phase.
\end{itemize}
There is a complementary way to see the failure that does not even appeal to the
horizon. Near the mediocrity saddle the overlap starts at \(m^0 \asymp d^{-1/2}\),
while the additive error floor in \eqref{app:intro:eq:veiga-bound} is already
\(\sqrt{\sfrac{\gamma}{pd}} \asymp d^{-1/2}\) at \(t = 0\). The uncertainty on
\(m\) is thus of the same order as \(m\) itself throughout the flat region: the
deterministic mean cannot resolve the sub-\(d^{-1/2}\) growth that carries the
student out of the saddle, and the exponential prefactor only compounds this.
\addnote{Equivalently, the bound controls a \emph{relative} error on \(m\) only
once \(m = \mathcal{O}(1)\); while \(m = \mathcal{O}(d^{-1/2})\) the signal and the
error bar coincide.}
This is the analytic content of the statement, made in
Section~\ref{sec:intro:exponents}, that in the search phase the fluctuations of
the order parameter cannot be neglected.

\paragraph{What replaces the deterministic argument.}
The deterministic ODE is therefore used in this thesis only where the horizon is
\(\mathcal{O}(1)\) or \(\mathcal{O}(\log d)\) --- the descent phase, and the
gradient-flow / mean-field / high-dimensional regimes of the phase diagram in
Figure~\ref{intro:fig:triangle}. The long search phase, which for \(\ell \geq 3\)
is the bottleneck that sets the sample complexity, is instead controlled by
\emph{separate stochastic arguments}, and Chapter~\ref{chap:paper-2305} is where
this is worked out in full and checked against simulation. There, the subsection
\emph{``Does stochasticity matter?''} (Section~\ref{2305:sec:sde_p1}) promotes the
overlap ODE to the effective diffusion of \citet{arous2022high} --- same drift,
fluctuation amplitude \(\sqrt{\sfrac{\gamma}{d}}\) --- and integrates it
\emph{through} the saddle, exactly where the bound \eqref{app:intro:eq:veiga-bound}
is vacuous. Solving the stochastic exit-time problem yields corrections of
relative order \(\sqrt{\sfrac{\gamma}{d}}\) that leave the deterministic exit-time
formulas \eqref{2305:eq:exit_time_p1}--\eqref{2305:eq:exit:quenched} unchanged at
leading order; and Figure~\ref{2305:fig:sde_p1} confirms numerically, at
\(d = 3000\), that SGD and the integrated SDE started from the same initialization
reach the signal at essentially the same time. Stochasticity is thus \emph{shown},
not merely assumed, not to alter the sample complexity.

The same conclusion is reached, by different means, in the chapters that need
polynomially long horizons directly: Chapters~\ref{chap:paper-2405}
and~\ref{chap:paper-2406} control the discrete process with high probability on the
random horizon \(t \leq T \wedge \tau_\eta\), through a martingale decomposition of
the drift, a discrete Gr\"onwall inequality, and sub-Gaussian concentration of the
per-step fluctuations --- none of which passes through
\eqref{app:intro:eq:veiga-bound}. All are instances of the information-exponent
analysis of \citet{arous2021online}: the deterministic reduction makes the
\emph{descent} phase tractable, while the \emph{search} phase belongs to the
stochastic theory.

\section{The partial-trace algorithm}
\label{app:intro:partial-trace}

This appendix describes the spectral algorithm of \citet{damian2024computational} that achieves weak recovery of the teacher direction \(\vec{w}^\star\) from \(n \asymp d^{\ell/2}\) samples, matching the CSQ lower bound discussed in Section~\ref{sec:intro:exponents}. We work in the single-index setting: \(\vec{x} \sim \mathcal{N}(0, I_d)\), \(\vec{w}^\star \in \mathbb{S}^{d-1}(\sqrt{d})\), \(\lambda^\star = \langle \vec{w}^\star, \vec{x}\rangle / \sqrt{d} \sim \mathcal{N}(0,1)\), and \(y = h^\star(\lambda^\star) = \sum_{k \geq \ell} a_k\,\mathrm{He}_k(\lambda^\star)\) with information exponent \(\ell\) and \(a_\ell \neq 0\). Let \(\hat{\vec{w}}^\star = \vec{w}^\star / \sqrt{d}\) denote the unit teacher direction.

\subsection{The population identity}

The algorithm rests on the family of multivariate Hermite tensors. For \(\vec{x} \sim \mathcal{N}(0, I_d)\) and \(\ell \geq 1\), the order-\(\ell\) Hermite tensor \(T_\ell(\vec{x}) \in (\mathbb{R}^d)^{\otimes \ell}\) is the unique degree-\(\ell\) symmetric tensor polynomial in \(\vec{x}\) orthogonal to all polynomials of lower degree in \(L^2(\mathcal{N}(0,I_d))\). The explicit component forms for small \(\ell\) are:
\begin{align*}
    T_1(\vec{x})_i =& x_i, \\
    T_2(\vec{x})_{ij} =& x_i x_j - \delta_{ij}, \\
    T_3(\vec{x})_{ijk} =& x_i x_j x_k
        - \bigl(\delta_{ij}\,x_k + \delta_{ik}\,x_j + \delta_{jk}\,x_i\bigr), \\
    T_4(\vec{x})_{ijkl} =& x_i x_j x_k x_l \\
        &- \bigl(\delta_{ij}\,x_k x_l + \delta_{ik}\,x_j x_l + \delta_{il}\,x_j x_k
               + \delta_{jk}\,x_i x_l + \delta_{jl}\,x_i x_k + \delta_{kl}\,x_i x_j\bigr) \\
        & + \bigl(\delta_{ij}\delta_{kl} + \delta_{ik}\delta_{jl} + \delta_{il}\delta_{jk}\bigr).
        \label{app:intro:eq:hermite-tensor-forms}
\end{align*}
Each formula is the outer product \(\vec{x}^{\otimes \ell}\) with the lower order corrections needed to enforce orthogonality: at order \(\ell = 2\) the identity is subtracted; at order \(\ell = 3\), the three symmetrized terms \(\delta_{ij}x_k + \delta_{ik}x_j + \delta_{jk}x_i\); at order \(\ell = 4\), six rank-two matrix terms and three rank-zero (double-\(\delta\)) corrections. The general construction follows the recurrence
\begin{equation}\label{app:intro:eq:hermite-tensor-recurrence}
    T_\ell(\vec{x}) = \operatorname{sym}\!\bigl(\vec{x} \otimes T_{\ell-1}(\vec{x})\bigr)
                    - (\ell-1)\,\operatorname{sym}\!\bigl(I \otimes T_{\ell-2}(\vec{x})\bigr),
    \quad T_0 = 1,\; T_1(\vec{x}) = \vec{x},
\end{equation}
where \(\operatorname{sym}\) denotes full symmetrization over all index permutations. This mirrors the scalar recurrence \(\mathrm{He}_\ell(t) = t\,\mathrm{He}_{\ell-1}(t) - (\ell-1)\,\mathrm{He}_{\ell-2}(t)\).

The defining property, which one verifies from~\eqref{app:intro:eq:hermite-tensor-recurrence} by induction, is the contraction identity: for any unit vector \(\vec{v} \in \mathbb{S}^{d-1}\),
\begin{equation}\label{app:intro:eq:hermite-tensor-contraction}
    T_\ell(\vec{x})[\underbrace{\vec{v},\,\ldots,\,\vec{v}}_\ell] = \mathrm{He}_\ell(\langle \vec{v},\,\vec{x}\rangle).
\end{equation}
For example, when \(\ell = 3\) this is immediate: contracting with \(\vec{v}^{\otimes 3}\),
\[
    T_3(\vec{x})[\vec{v},\vec{v},\vec{v}]
    = (\langle\vec{v},\vec{x}\rangle)^3
      - \underbrace{\bigl(\delta_{ij}x_k + \delta_{ik}x_j + \delta_{jk}x_i\bigr)v_iv_jv_k}_{=\;3\,\|\vec{v}\|^2\langle\vec{v},\vec{x}\rangle\;=\;3\langle\vec{v},\vec{x}\rangle}
    = \mathrm{He}_3(\langle\vec{v},\vec{x}\rangle),
\]
since each term \(\delta_{ij}x_k v_iv_jv_k = \|\vec{v}\|^2\langle\vec{v},\vec{x}\rangle = \langle\vec{v},\vec{x}\rangle\).

Setting \(\vec{v} = \hat{\vec{w}}^\star\) in~\eqref{app:intro:eq:hermite-tensor-contraction} returns \(\mathrm{He}_\ell(\lambda^\star)\). The Hermite expansion \(h^\star(\lambda^\star) = \sum_{k \geq \ell} a_k\,\mathrm{He}_k(\lambda^\star)\) therefore rewrites as a series of fully contracted tensors:
\begin{equation}\label{app:intro:eq:hstar-tensor-expansion}
    h^\star(\lambda^\star)
    = \sum_{k \geq \ell} a_k\;T_k(\vec{x})\bigl[\underbrace{\hat{\vec{w}}^\star,\ldots,\hat{\vec{w}}^\star}_k\bigr].
\end{equation}
The \emph{population identity} follows from tensor Hermite orthogonality. For each \(k\), taking the expectation of~\eqref{app:intro:eq:hstar-tensor-expansion} multiplied by \(T_\ell(\vec{x})\):
\[
    \mathbb{E}\!\left[T_k(\vec{x})\bigl[(\hat{\vec{w}}^\star)^{\otimes k}\bigr] \cdot T_\ell(\vec{x})\right]
    = \mathbb{E}\!\left[\mathrm{He}_k(\lambda^\star)\cdot T_\ell(\vec{x})\right].
\]
For \(k \neq \ell\) this vanishes because tensors of different degree are orthogonal in \(L^2(\mathcal{N}(0,I_d))\). For \(k = \ell\), the second-moment identity
\(\mathbb{E}[(T_\ell(\vec{x}))_{i_1\cdots i_\ell}(T_\ell(\vec{x}))_{j_1\cdots j_\ell}] = \ell!\,\delta^{\mathrm{sym}}_{(i_1\cdots i_\ell)(j_1\cdots j_\ell)}\),
with \(\delta^{\mathrm{sym}}\) the symmetrized Kronecker product, gives after contracting \(\ell\) indices against \(\hat{\vec{w}}^\star\):
\[
    \mathbb{E}\!\left[\mathrm{He}_\ell(\lambda^\star)\cdot T_\ell(\vec{x})_{j_1\cdots j_\ell}\right]
    = \ell!\,\hat{w}^\star_{j_1}\cdots\hat{w}^\star_{j_\ell}.
\]
Summing over \(k\), and using \(a_k = 0\) for \(k < \ell\) by definition of the information exponent, the series collapses to the single term \(k = \ell\):
\begin{equation}\label{app:intro:eq:pop-identity}
    \mathbb{E}\!\left[\,y\cdot T_\ell(\vec{x})\,\right] = a_\ell\cdot\ell!\cdot(\hat{\vec{w}}^\star)^{\otimes \ell}.
\end{equation}
The right-hand side is a rank one spike of \(\Theta(1)\) magnitude pointing exactly in the teacher direction. Observing this tensor directly would achieve recovery, but it has \(d^\ell\) entries and we have only \(n\) samples. The partial trace collapses it to a matrix or a vector while preserving the rank one structure.

\subsection{The estimator}

Form the empirical version of Equation~\eqref{app:intro:eq:pop-identity} and contract identity pairs to reduce the order of the tensor. Write \([\,I^{\otimes r}\,]\) for the operation of contracting \(r\) pairs of free indices against the \(d \times d\) identity.

\medskip\noindent
\textbf{\(\ell\) even.} Contract \((\ell-2)/2\) index pairs, landing on a \(d\times d\) matrix:
\begin{equation}\label{app:intro:eq:estimator-even}
    M_n = \frac{1}{n}\sum_{i=1}^n y_i\cdot T_\ell(\vec{x}_i)\!\left[\,I^{\otimes(\ell-2)/2}\,\right] \;\in\; \mathbb{R}^{d\times d}, \qquad \hat{\vec{w}} = v_1(M_n),
\end{equation}
where \(v_1(\cdot)\) denotes the leading eigenvector.

\medskip\noindent
\textbf{\(\ell\) odd.} Contract \((\ell-1)/2\) index pairs, landing on a vector:
\begin{equation}\label{app:intro:eq:estimator-odd}
    \vec{v}_n = \frac{1}{n}\sum_{i=1}^n y_i\cdot T_\ell(\vec{x}_i)\!\left[\,I^{\otimes(\ell-1)/2}\,\right] \;\in\; \mathbb{R}^d, \qquad \hat{\vec{w}} = \frac{\vec{v}_n}{\norm{\vec{v}_n}}.
\end{equation}

\medskip
Taking the expectation in~\eqref{app:intro:eq:estimator-even} and using~\eqref{app:intro:eq:pop-identity}: each traced identity pair uses \(\langle\hat{\vec{w}}^\star,\hat{\vec{w}}^\star\rangle = 1\), so contracting \((\hat{\vec{w}}^\star)^{\otimes\ell}\) against \((\ell-2)/2\) identity matrices leaves a rank one matrix:
\begin{equation}\label{app:intro:eq:pop-mean}
    \mathbb{E}[M_n] = a_\ell\cdot\ell!\cdot\hat{\vec{w}}^\star(\hat{\vec{w}}^\star)^\top
    = \frac{a_\ell\cdot\ell!}{d}\cdot\vec{w}^\star(\vec{w}^\star)^\top,
\end{equation}
a rank one spike of strength \(a_\ell\cdot\ell! = \Theta(1)\) along \(\vec{w}^\star\). In the odd case, \(\mathbb{E}[\vec{v}_n] = a_\ell\cdot\ell!\cdot\hat{\vec{w}}^\star\), a vector of \(\Theta(1)\) norm aligned with \(\vec{w}^\star\).

\subsection{Why \(n \asymp d^{\ell/2}\) is the threshold}

Consider the even case; the odd case is analogous. Decompose \(M_n = \mathbb{E}[M_n] + (M_n - \mathbb{E}[M_n])\).

\medskip\noindent\textbf{Signal.} From~\eqref{app:intro:eq:pop-mean}, the spike has operator norm \(\norm{\mathbb{E}[M_n]}_\mathrm{op} = \Theta(1)\).

\medskip\noindent\textbf{Noise.} The fluctuation \(M_n - \mathbb{E}[M_n]\) is an average of \(n\) i.i.d.\ mean zero matrices. Each summand \(y_i\cdot T_\ell(\vec{x}_i)[\,I^{\otimes(\ell-2)/2}\,]\) has \(O(1)\) entries, but a large operator norm: contracting the order-\(\ell\) Hermite tensor down to a matrix leaves a residual operator norm growth of \(d^{(\ell-2)/2}\), which combined with the matrix dimension \(d\) places the fluctuation scale at
\[
    \norm{M_n - \mathbb{E}[M_n]}_\mathrm{op} \;\asymp\; \sqrt{\frac{d^{\ell/2}}{n}}\cdot\sqrt{\mathbb{E}[h^{\star 2}]}.
\]
This is the heavy-tailed noise phenomenon: unlike standard PCA, where the contracted noise has operator norm \(d^{1/2}\), the contracted degree-\(\ell\) Hermite tensor yields operator norm \(d^{\ell/2}\).

\medskip\noindent\textbf{Detectability.} The spike is visible above the bulk if and only if signal \(\gtrsim\) noise:
\begin{equation}\label{app:intro:eq:bbp}
    \Theta(1) \;\gtrsim\; \sqrt{\frac{d^{\ell/2}}{n}} \qquad\Longleftrightarrow\qquad \boxed{n \;\gtrsim\; d^{\ell/2}.}
\end{equation}
Above this threshold the leading eigenvector of \(M_n\) satisfies \(\langle \hat{\vec{w}},\,\hat{\vec{w}}^\star\rangle \geq \mathrm{const} > 0\), achieving weak recovery. Below it the spike is submerged in the semicircular bulk (a BBP-type transition at \(n \asymp d^{\ell/2}\)) and \(\lvert\hat{\vec{w}}\cdot\hat{\vec{w}}^\star\rvert \approx d^{-1/2}\), no better than a random initialization.

\subsection{Familiar special cases}

\(\ell = 2\): The contraction \([\,I^{\otimes 0}\,]\) is trivial, and \(T_2(\vec{x}) = \vec{x}\vec{x}^\top - I\), so the estimator reduces to the standard PCA matrix,
\[
    M_n = \frac{1}{n}\sum_{i=1}^n y_i\!\left(\vec{x}_i\vec{x}_i^\top - I\right), \qquad \mathbb{E}[M_n] = 2\,a_2\,\hat{\vec{w}}^\star(\hat{\vec{w}}^\star)^\top,
\]
recovering \(\vec{w}^\star\) at \(n \asymp d\).

\(\ell = 3\): The vector estimator \(\vec{v}_n = \frac{1}{n}\sum_i y_i\,T_3(\vec{x}_i)[\,I\,]\) contracts one index pair of the third-order tensor to a vector aligned with \(\hat{\vec{w}}^\star\), recovering \(\vec{w}^\star\) at \(n \asymp d^{3/2}\).
