\chapter{Introduction}
\label{chap:introduction}

\section{Motivation}
\begin{itemize}
    \item AI is the word of this decade. Since the release of ChatGPT, there has been an explosion of interest in AI and machine learning. This unforeseen interest for a new technology has been possible mostly because an astonishing engeenering challange has been solved: we have created a pipeline that is basically able to digest the entire human digital knowledge and use it to train deep learning models. This a huge achivement involeving many different fields, from hardware design to software engineering, and it tracks back to the first attempt to replicate the human brain with artificial neural networks in the 1950s, through the development of backpropagation in the 1980s, to the advent of GPU computing in the 2000s, and the rise of big data in the 2010s. However, while we have made tremendous progress in building and training deep learning models, our theoretical understanding of how these models work is still limited.
    \item Understanding the theoretical foundations of deep learning is crucial for several reasons. The most direct one are practical: it can help us design better algorithms, improve the performance of existing models (saving resources and energy). The second it might be social: it can help us understand the limitations and potential risks of AI, and develop strategies to mitigate them: there is the full line of research of interpretability, fairness, robustness. But the one this is a thesis in physics, and as physicist we are not driven (only) by practical applications, but also by the desire to understand the fundamental principles that govern the behavior of complex systems. Satisfy human curiosity is a noble goal in itself, and it can lead to unexpected discoveries and insights that can have far-reaching implications beyond the original motivation. We are building a system that will be surrounding us, affecting most of the aspects of our life, it is natural human behavior to want to understand how it works, and what are the principles that govern its behavior.
    \item Many machine learning model are theoretically grounded by construction. For example, support vector machines are based on the principle of maximizing the margin between classes, and their theoretical properties can be analyzed using tools from convex optimization and statistical learning theory. Similarly, decision trees are based on the principle of recursive partitioning, and their theoretical properties can be analyzed using tools from combinatorics and probability theory. However, deep learning models have always been difficult to understand. The core idea of using an big number of parameters to fit data breaks many of the assumptions that underlie traditional statistics, bias-variance tradeoff to name one. Deep learning showed that the model can learn effective representations of data, and generalize well to unseen examples, embedded in a non-convex optimization landscape, and trained with simple algorithms like stochastic gradient descent. This has challenged many of the assumptions and intuitions that we had about learning and generalization, and has led to a lot of confusion and debate in the research community.
    % Why we do what we do
    \item Despite what you could think looking at its content, this is a thesis submitted to the \emph{Ècole Doctorale de Physique}(EDPY). Therefore we should at least have a physicist approach to the subject. This means that we should not be afraid to make assumptions, and to build our model of deep learning.
    \item We are building simple toy-model because that's where you start. If you want ti understand how the mechanics of the universe work you start from  toy model of the atom, not from the full quantum field theory. In the same way, we should always be looking at experimental data, and be ready
    to stir the theory if the data does not fit. We are already lucky to be able to run experiments in our computers, and do not need crazy expensive particle colliders to do so. So we should take advantage of that! Moreover, another advantage of natural science is that the fundamental process at the core is well definied mathematically: the optimization process of training a neural network is a well defined mathematical process, we do not need to figure out the \emph{what} of the process, but only the \emph{why} and the \emph{how}.
    \item Think of where chemistry was at the end of the 19th century: practitioners could synthesize remarkably complex molecules and predict reactions, but nobody understood the deep mechanism, why atoms bond the way they do. That's roughly where we are. We're trying to build the Thomson model: the first crude, probably-wrong picture of the underlying structure. It won't be correct, but it'll be concrete enough to be tested and improved. We want to be Thomson, so that someone can be Bohr after us, and eventually someone gets to quantum mechanics.
    \item We want a model simple enough to be analytically tractable, but complex enough to show interesting phenomena appearing at large scale. We want to have a theory AND experiments supporting it.
    \item As of the many aspect of deep learning that are not well understood, we focus on dynamical process of training that takes ransom parameters to working ones. The loss landscape of deep learning is highly non-convex, and it is not clear why simple optimization algorithms like stochastic gradient descent are able to find good solutions. The dynamics induced by SGD brings in a bias that correlates with generalization. We focus on this
    % Probably subsection: Past attempts to understand deep learning
    \item Classic learning theory (PAC learning, VC dimension, etc.) focuses on the worst-case scenario. This can lead to pessimistic bounds that do not reflect the performance of learning algorithms in practice. Statistical physic learning theory, on the other hand, provides a framework for analyzing the typical-case performance of learning algorithms, which can lead to more realistic bounds and insights.
    \item This connection between physics (statistical physics in particular) and machine learning is not new at all. First it has been used to study capacity of neural networks, and the generalization properties of learning algorithms. In the end machine learning is a problem of handling big amount of data, with a big amount of parameters, and this is exactly the kind of problem that statistical physics is designed to handle. 
\end{itemize}

\section{Our toy model and terminology}
\begin{itemize}
    \item We go with supervised learning. 
    \item For a machine learning algorithm we need 3 ingredients: a data distribution, a model class, and a learning algorithm. All the 3 have enormous freedom, and we can choose them in a way that is both simple and interesting. As a good physicist, we will make many assumptions, trying to keep only the essential ingredients.
    \item Our minimum requirement is to have a non-convex optimization landscape in the space of parameters, an architecture that resemble the core of deep learning, building feature by non-linearly combine features from the previous layer, and a learning algorithm that is a gradient methos as close as possible to the one used in practice.
    \item This are kind of the axioms of our theory. For sure they are restrictive, and they do not capture all the complexity of deep learning, but there is a lot to be understood even in this simplified setting.
    % The data
    \item The distribution of our data is gaussian. The labels are generated by a model that depens only on a fixed and finate subspace of the input: $y = f(x \cdot W^\star)$
    \item Gaussian data has not structure and this is indeed a limitation. We need because the gaussian structures propagates nice properties through the network, and it makes the model analytically tractable. For examples preactivations (or local fields how we could call them borrowing from statistical physics) $z = W \cdot x$ are also gaussian.
    \item Attempts to inquiry the gaussianity of the assumptions has been done (here I should discuss and cite universality).
    % The model
    \item We study a two-layer neural network, expressible this way: $f(x) = \sum_{i=1}^N a_i \sigma(w_i \cdot x)$. This is the simplest model that can show the phenomenon we want to study, and it is already non-convex.
    \item Note that the model is a special case of our data generator model, This is why we often refer as teacher student setting.
    % The algorithm
    \item We study online batched SGD (even though often batch size is 1). We initialize gaussianly. This is not what it is used in practice, where more fancy mini-natch multi-epoch algorithms are used, with fancy optimizers like Adam or more recently Muon. Traditional supervised learning usually repeat the data across epochs, but in the online setting we just keep feeding the model with new data. Modern deep learning where the data is do big that it is not the bottleneck, practionioers also train in an online fashion; but their samples are obviously not independent like ours, since they have the structural correlation of the data. We do it because we want independence between steps, because it makes object more tractable. % Remeber to make a margin note saying that we drop this assumption in Repetita Iuvant chapter, and this was actualy a pivotal moment for the development of the theory.
\end{itemize}


