\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

\begin{noparskip}

Understanding why gradient-trained neural networks succeed requires theoretical models that capture how they discover low-dimensional structure buried in high-dimensional data. Despite the empirical dominance of large-scale deep learning, even small, shallow networks remain analytically elusive: the dynamics that govern what they learn, and how fast, are not yet fully understood.
This thesis builds on a statistical-physics inspired  theory of these dynamics: studying stochastic gradient descent on shallow networks and simplified attention architectures trained on synthetic, teacher-generated data, by tracking the evolution of a handful of summary statistics rather than every individual weight.
In the high-dimensional limit, this evolution is described exactly by a closed system of deterministic differential equations, making the speed and difficulty of learning fully quantifiable.

Initially, we frame previously separate descriptions of this dynamics --- the classical gradient-flow, high-dimensional, and overparameterized mean-field regimes --- as special cases of a single underlying process. Within this framework, we give closed-form sample-complexity formulas for learning hard, phase-retrieval-type targets, and show that increasing network width speeds up learning only by a constant factor, without escaping the hardness controlled by the target's information exponent.

The second part studies how simple modifications to vanilla stochastic gradient descent move training out of this hardness landscape. A small amount of data reuse, seeing each sample twice rather than once, shifts the obstruction away from the \emph{information exponent} that governs one-pass learning, to a smaller \emph{generative exponent}, recovering almost any hidden direction efficiently, with a hierarchical mechanism accounting for the remaining hard cases. We identify the batch size that minimizes training time for a target of given difficulty, and introduce a modified loss that removes the resulting speed limit altogether.

Finally, we extend the analysis beyond static inputs to sequential data and a simplified attention layer, showing that positional structure can substantially accelerate learning relative to unstructured inputs.

Together, these results give a solvable, unified theory of the training dynamics of shallow networks and simple attention models. They quantify when gradient-based learning succeeds at uncovering hidden structure, how far simple algorithmic changes can push that success, and which properties of the data and training govern its speed.

\textbf{Keywords:} deep learning, neural networks, stochastic gradient descent, training dynamics, statistical physics, mean-field theory, teacher-student models, sample complexity, phase retrieval, attention mechanisms
\end{noparskip}
